---
title: "Homework Assignment 1"
author: "Ellen Tomljanovic & Bryce Garver"
date: 'Due EoD Sun Oct 31, 2021'
output:
  html_document: 
    toc: TRUE
    toc_float: TRUE
    code_folding: show
  pdf_document: default
---

## HW-1 Submission

## Data frame basics
```{r, cache = TRUE}
#Reading & Exploring Data
nyc <- read.csv("nycflights.csv")
head(nyc,n=10)
str(nyc)
summary(nyc)

```
## Cleaning Data
```{r, cache = TRUE}
#Remove Unneeded Columns
nyc$X <- NULL
nyc$year <- NULL

#Change Column Type
nyc$carrier <- as.factor(nyc$carrier)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$origin <- as.factor(nyc$origin)
nyc$flight <- as.factor(nyc$flight)
nyc$dest <- as.factor(nyc$dest)
nyc$month <- as.factor(nyc$month)

#Additional Cleaning
nyc$flightdate <- as.Date(nyc$time_hour, format = "%Y-%m-%d %H:%M:%S")
str(nyc)
```


## Data Exploration
```{r, cache = TRUE}

#How many airlines are there?
nlevels(nyc$carrier)

#How many flights there were by the airline with code `OO`?
nrow(nyc[nyc$carrier == "OO",])

#How long is the shortest flight out of any NYC airport?
min(nyc$air_time, na.rm = TRUE)

#How many flights where there by United Airlines on Jan 12th 2013?
nrow(nyc[nyc$carrier == "UA" & nyc$flightdate == "2013-01-12",])

```

## Arrival Delay
```{r}
#What was the average arrival delay for all airports and all airlines combined in 2013? 
mean(nyc$arr_delay, na.rm = TRUE) 

#Whats was the median arrival delay for all airports and all airlines combined in 2013?
median(nyc$arr_delay, na.rm = TRUE) 

```
It's very interesting that the mean is positive (meaning flights arrive late) and that the median is negative. This would say that there are a handful of flights with significant delays that are skewing the mean to be much higher. It may make sense to remove them as outliers if we were looking for a more accurate depiction of on average how late/early are flights

### Airline Performance
```{r}
#Calculate average arrival delays by airline.
carrier_delay_mean <-round(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE))
barplot(sort(carrier_delay_mean), col = "blue")

names(carrier_delay_mean[carrier_delay_mean == max(carrier_delay_mean)])
names(carrier_delay_mean[carrier_delay_mean == min(carrier_delay_mean)])

names(carrier_delay_mean[carrier_delay_mean < 0])


```
The airline with the highest average arrival delay is `r names(carrier_delay_mean[carrier_delay_mean == max(carrier_delay_mean)])`. The airline with the lowest average arrival delay is `r names(carrier_delay_mean[carrier_delay_mean == min(carrier_delay_mean)])`. The airlines with negative average delay are `r names(carrier_delay_mean[carrier_delay_mean < 0])`.


### Air Gain
```{r}
nyc$airgain <- nyc$dep_delay - nyc$arr_delay
boxplot(airgain~carrier, data=nyc, col="blue")

tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE)


```
Most airlines, on average, do gain time back when in air. `r tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE)`. The airlines with the worst airgain are F9, FL, and MQ. The best appear to be HA, VX, and AS. 



